Cognate beginnings to bilingual lexical acquisition

### 2023-07-23 manuscript/manuscript.qmd
Simplify dependencies

---
abstract: |
    Bilingual infants show equivalent developmental trajectories of lexical acquisition compared to their monolingual peers. This is remarkable, given the increased complexity of their linguistic input. Recent studies suggested that bilingual vocabulary growth is boosted by the number of cognates shared by the pair of languages being learned, and that this facilitation effect is driven by a stronger parallel activation of cognates during linguistic exposure, compared to non-cognates. The mechanisms behind this facilitation are still unclear. In this study, we capitalise on accumulator models of language acquisition to propose an account of bilingual lexical acquisition in which parallel activation increases the rate at which children accumulate learning instances for words in both languages, even in fully monolingual situations. Under this hypothesis, we predicted a stronger cognate facilitation for words to which children were exposed less frequently (low-exposure words), as they are co-activated by their translation more often than high-exposure words do. We developed an extensive online vocabulary checklist, the Barcelona Vocabulary questionnaire (BVQ) to collect vocabulary data from 366 Catalan-Spanish bilingual toddlers aged 12 to 32 months. We then used Bayesian explanatory item response theory to model the acquisition trajectories of 302 Catalan/Spanish translation equivalents. We found an interaction between exposure and cognateness, which pointed to cognateness facilitating the aquisition of low-exposure words, but not of mean exposure or high-exposure words. Overall, our findings suggest that cognateness plays a key role in bilingual lexical acquisition, and provides evidence for a frequency-mediated facilitation effect driven by parallel activation.

thanks: |
    The authors declare no conflicts of interest with regard to the funding source of this study. GGC was supported by the a FPI 2020 research contract (PRE2019-088165) in a project funded by the Spanish Ministry of Science, Innovation, and Universities (PGC2018-101831-B-I00). DAV was supported by the European Union's Horizon 2023 research and innovation program under Marie Skłodowska–Curie Grant (765556) and a postdoctoral fellowship from the Foundation for Science and Technology of Portugal (UIDB/00214/2020). NSG was supported by Academia 2019 award from the Catalan Institution for Research and Advanced Studies (ICREA). We are grateful to Chiara Santolin, Ege E. Özer, Alicia Franco-Martínez, Cristina Rodríguez-Prada, and the rest of the Speech Acquisition and Perception research group for their helpful feedback. We thank Xavier Mayoral, Silvia Blanch, and Cristina Cuadrado for their technical support, and Cristina Dominguez and Katia Pistrin for their efforts in recruiting infants. We also thank all families and infants who participated in the experiments.
---

@@

@@

The foundations of word learning are in place early in age. At six months, infants start directing their gaze to some objects when hearing their labels [@tincoff1999beginnings; @bergelson2012months; @bergelson2015early], and shortly after caregivers start reporting some words as acquired by their infant in vocabulary checklists [e.g., @fenson2007macarthurbates; @samuelson2021precision]. Most research on early word acquisition relies extensively on data from monolingual children, and is oblivious to the fact that a substantial proportion of the world population acquires more than one language from early ages [@grosjean2021extent]. Previous work on bilingual vocabulary acquisition pointed to bilingual toddlers knowing, on average, less words in each of their languages than their monolinguals peers, and to both groups knowing a similar number of words---if not more words---when the bilinguals’ two languages are pooled together. @hoff2012dual found that English-Spanish bilingual toddlers in South Florida (United States) knew less words in English than monolinguals did, but both groups knew a similar total amount of words when both English and Spanish vocabularies were counted together. Other studies have provided converging evidence that bilinguals know a similar or even larger number of words than monolinguals when the two languages are aggregated [@oller2002language; @pearson1994patterns; @pearson1993lexical; @patterson2004comparing; @patterson2004bilingual; @smithson2014bilingualism; @petitto2001bilingual; @gonzalez-barrero2020bilingual]. A more detailed analysis of the words in bilinguals’ lexicon show some interesting patterns.

One important observation of studies on bilinguals' early vocabulary acquisition is that cognate words are easier to acquire than non-cognate words. Cognate words are translation equivalents that are phonologically similar (or share some type of form-similarity). For instance, the Spanish translation equivalent of *cat* is *gato*, a cognate word; the translation equivalent of *dog* is *perro*, a non-cognate word. The differences in the percentages of cognate and non-cognate words between two languages is related to historic reasons: languages typologically close (like Dutch and English or Italian and Spanish) share more cognates than languages typologically distant (like English and Chinese, or Urdu and Spanish). The conclusion that cognate words are easier to learn is based on two types of evidence: studies investigating vocabulary sizes in children learning language pairs with different percentages of cognates (that is, differing in their typological distance) and studies comparing the number of cognate and non-cognate words children know in a specific language pair.

@floccia2018introduction published an impressive study comparing vocabularies of children learning several language pairs differing in their percentage of cognates. The authors collected vocabulary data on word comprehension and production from 372 24-month-old bilingual toddlers living in the United Kingdom who were learning English and an additional language. The additional language was one of 13 typologically diverse languages: Bengali, Cantonese Chinese, Dutch, French, German, Greek, Hindi/Urdu, Italian, Mandarin Chinese, Polish, Portuguese, Spanish, and Welsh. The authors calculated the average phonological overlap between the words in each of these additional languages and their translation equivalents in English, which was taken as a proxy of the degree of cognateness between each pair of languages. Floccia and co-workers reported an increase in vocabulary size in the additional language (i.e., not English) associated with an increase in the average phonological similarity between the translation equivalents of each language pair. For example, English-Dutch bilinguals (languages with a high phonological overlap), were able to produce more Dutch words than English-Mandarin bilinguals (languages with a low phonological overlap) were able to produce in Mandarin. @blom2020crosslanguage, @bosma2019longitudinal, and @gampe2021does reported similar results, providing converging evidence of a facilitatory effect of a lower language distance (i.e., higher degre of cognateness) on vocabulary size.

A second set of studies concluded that cognates are overrepresented in bilinguals’ early lexicon. @bosch2014first collected parental reports of expressive vocabulary from 48 Catalan-Spanish bilinguals aged 18 months and found that cognates represented a larger proportion of participant’s vocabulary than non-cognates. @schelletter2002effect provided converging evidence from a longitudinal single-case study, in which an English-German bilingual child produced cognates earlier than non-cognates, on average. @mitchell2022cognates addressed this issue in a larger sample longitudinal study. These authors collected expressive vocabulary data of 47 16- to 30-month-old French-English bilinguals living in Canada, in both languages. They created two lists of translation equivalents: one made of 131 cognates, and one made of 406 non-cognates. The proportion of words that children were reportedly able to produce was higher in the cognate lists than in the non-cognate list across ages, even when both lists were matched by semantic category (furniture, animals, food were similarity represented in both lists) and age-of-acquisition norms (an index of word difficulty). Taken together, the results of these two lines of research provide a strong support to the hypothesis that phonological similarity (as reflected in cognateness) plays a facilitation role in bilingual word acquisition.

@@

We propose an account in which a learning instance for a word may also represent a learning instance for its translation equivalent, to the extent that such translation equivalent is co-activated^[We use the term *learning instance* in the fashion of accumulator models of language acquisition; as an exposure to a word-form that constitutes an opportunity for the child to accumulate information about the word. We do not make strong assumptions about whether a learning instance is a discrete or a continuous unit of accumulation of information. We rather consider that a learning instance of a word is so to the degree of the strength of activation of its phonological representation. This activation may result from the infant being exposed to the word-form, or the result of activation spreading through phonological or semantic links across lexical representation, as in the case of parallel activation.]. The strength of this co-activation is proportional to the phonological similarity between the two translation equivalents; given that cognates share higher phonological similarity than non-cognates, the former should be co-activated more frequently than the latter. This should lead to a faster accumulation of learning instances for cognates, compared to non-cognates. Parallel activation would allow bilingual children to accumulate learning instances for words in both languages even during fully monolingual situations, but the impact of this mechanism would be asymmetric across languages: words from the lower-exposure language (i.e., the language to which the infants are exposed to more often) would receive additional activation through parallel activation more frequently than words from the higher-exposure language. Therefore, the acquisition of words from the lower-exposure language would benefit more strongly from their cognate status than words from the higher-exposure language.

@@

n_item_language <- bvq_data$pool |> 
    count(language) |>
    pull(n) |>
    set_names(c("catalan", "spanish"))

n_items_version <- bvq_data$pool |>
    unnest(version) |> 
    count(version, language) |> 
    pivot_wider(names_from = language,
                values_from = n) |> 
    janitor::clean_names() |> 
    mutate(version = tolower(version))

@@

tbl_data <- items |> 
    distinct(te, .keep_all = TRUE) |> 
    arrange(language, list) |> 
    mutate(a = grepl("A", list),
           b = grepl("B", list),
           c = grepl("C", list),
           d = grepl("D", list)) |> 
    summarise(across(a:d, sum), 
              items = list(meaning),
              .by = semantic_category) |> 
    mutate(
        total = rowSums(across(a:d)),
        items = map(items, function(x) {
            paste(tail(x, 2),
                  collapse = ", ")
        })) |>  
    unnest_longer(items) |> 
    arrange(desc(total)) |> 
    relocate(semantic_category, a:d, total, items) |> 
    select(-total)

@@

knitr::kable(tbl_data,
             format = "latex", 
             digits = 1, 
             booktabs = TRUE,
             col.names = col_names,
             align = "lccccl")  |> 
    kableExtra::row_spec(nrow(tbl_data)-1, 
                         hline_after = TRUE)
```

@@

suppressWarnings({
    diff_responses <- responses |>
        distinct(id, age) |>
        left_join(distinct(participants, id, time_stamp, age),
                  by = c("id", "age")) |> 
        group_by(id) |> 
        arrange(id, time_stamp) |> 
        mutate(diff_lag = time_stamp-dplyr::lag(time_stamp, default = NA)) |>
        summarise(diff_min = min(diff_lag, na.rm = TRUE),
                  diff_max = max(diff_lag, na.rm = TRUE),
                  .groups = "drop") |> 
        dplyr::filter(!(diff_min %in% c(-Inf, Inf))) |>
        summarise(diff_min = min(diff_min, na.rm = TRUE),
                  diff_max = max(diff_max, na.rm = TRUE))
})

@@

dates <- participants$time_stamp |> 
    range(na.rm = TRUE) |> 
    lubridate::as_date()

@@

options(knitr.kable.NA = 0)

tbl_data <- participants |>
    mutate_at(vars(starts_with("doe_")), 
              (\(x) floor(x * 10) / 10)) |> 
    mutate(
        age = cut(floor(age),
                  c(10, 14, 18, 22, 26, 30, 34, 36),
                  include.lowest = TRUE),
        doe_catalan = cut(
            doe_catalan,
            c(0, 0.25, 0.5, 0.75, 1),
            labels = c("0-25%",
                       "25-50%", 
                       "50-75%",
                       "75-100%"),
            include.lowest = TRUE
        ) |>
            as.character()
    ) |>
    count(age, doe_catalan) |>
    arrange(desc(doe_catalan)) |>
    pivot_wider(names_from = age, 
                values_from = n,
                names_repair = janitor::make_clean_names) |> 
    mutate(across(where(is.numeric), \(x) ifelse(is.na(x), 0, x)))

sum_rows <- function(x) if (is.numeric(x)) sum(x) else "N"
sums_row <- as.data.frame(lapply(tbl_data, sum_rows),
                          col.names = colnames(tbl_data))
tbl_data <- rbind(tbl_data, sums_row)

knitr::kable(tbl_data,
             format = "latex", 
             digits = 1, 
             booktabs = TRUE,
             col.names = col_names,
             align = "lcccccc")  |> 
    kableExtra::add_header_above(c(" " = 1,
                                   "Age (months)" = ncol(tbl_data)-1)) |>
    kableExtra::row_spec(nrow(tbl_data)-1, 
                         hline_after = TRUE)
```

@@

n_items_final <- n_items_duplicated |> 
    count(language) |> 
    group_split(language) |> 
    set_names(c("Catalan", "Spanish")) |>
    map_int("n")

@@

rope_interval <- scales::percent(c(lower = -0.1/4,
                                   upper = 0.1/4),
                                 accuracy = 0.1)

@@

We developed the *Exposure* predictor to account for the fact that bilinguals' exposure to a given word-form is not only a function of the word-forms lexical frequency, but also of the quantitative input they receive from the language such word-form belongs to, we expressed lexical frequencies as the product between both variables. First, we extracted the child-directed lexical frequency of each word from the CHILDES database [@macwhinney2000childes; @sanchez2019childesdb]. Using the corresponding lexical frequencies directly from Catalan and Spanish was not possible due to the low number of Catalan participants and tokens available in their corresponding CHILDES corpora, so they were extracted from the English corpora instead. We then mapped the lexical frequencies of the English words to their Catalan and Spanish translations [see @fourtassi2020growth for a similar approach], and transformed them to Zipf scores [@vanheuven2014subtlexuk; @zipf1949human]. We then multiplied the resulting lexical frequencies by the reported degree of exposure of the child to Catalan or Spanish. For instance, for a child whose degree of exposure is 80% for Catalan and 20% for Spanish, the expected *Exposure* score to the Catalan word *cotxe* [*car*]---with a lexical frequency of `r round(items$freq[items$item=="cotxe"], 2)`---would be `r round(items$freq[items$item=="cotxe"]*0.80, 2)`, while that of its translation to Spanish *coche* would be `r round(items$freq[items$item=="coche"]*0.20, 2)`.

We defined *Cognateness* as the phonological similarity between each word-form and its translation. For each translation equivalent, we used the `stringdist` [@vanderloo2014stringdist] R package to calculate the Levenshtein distance between the Catalan and the Spanish phonological transcriptions of the word-forms. The Levenshtein distance measures the number of editions (insertions, deletions, or substitutions) that one string of phonemes/characters must go through to become identical to the other [@levenshtein1966binary]. We divided the Levenshtein distance of each translation equivalent by the length of the longest word-form to correct for word length (longer strings are likely to show a larger number of mismatches). Finally, we subtracted the result from one so that it could be interpreted in terms of phonological similarity, instead of phonological distance. This led to a distance metric that ranged from zero to one, where zero indicates that both word-forms are completely different (e.g., /\textipa{"taw.l5}-\textipa{"me.sa}/, *table*), and one indicates that the two word-forms are identical (e.g., /\textipa{"mar}-\textipa{"mar}/, *sea*) [see @floccia2018introduction; @heeringa2003norwegian; @schepens2012distributions; @fourtassi2020growth; @laing2022phonological for similar approaches].

@@

We implemented the model using `brms` [@burkner2017brms], a R interface to the Stan probabilistic language (`r stan_version`) [@carpenter2017stan]. We ran four iteration chains using the by-default No U-Turn Sampler algorithm with `r format(niterations(model_fit), big.mark = ",")` iterations each and an additional `r format(niterations(model_fit), big.mark = ",")` warm-up iterations per chain. Model posterior draws and predictions were handled using the `tidybayes` [@kay2021tidybayes] and `marginaleffects` [@arel-bundock2022marginaleffects] R packages.

@@

post_draws_list <- model_summary |> 
    mutate(across(.median:.lower,
                  \(x) ifelse(grepl("Intercept", .type),
                              plogis(x), x/4))) |> 
    split(model_summary$.variable)
```

@@

```{r tbl-coefs}
#| label: tbl-coefs
#| tbl-cap: "Posterior distribution of regression coefficients. Median: median of the posterior distribution in the probability scale. 95\\% HDI: 95\\% highest density interval of the distribution. p(ROPE): overlap between the 95\\% HDI and the ROPE, indicating the posterior probability that the true value of the coefficient is equivalent to zero. In the right column, we show a graphical representation of the median and 95% HDI of each coefficient, with a reference dotted line indicating the location of zero."
#| echo: false
#| message: false
#| warning: false
#| out-width: 80%
# summarise posterior draws
intercept_label <- "Intercepts (at 22 months)"

tbl_data <- model_summary |>
    select(.variable_name, .median, .lower, .upper, .rope_overlap, .type) |>
    mutate(
        across(.median:.upper,
               \(x) ifelse(.type==intercept_label, plogis(x), x/4)),
        .variable_name = gsub("×", '$\\\\times$', .variable_name),
        .hdi = paste0("[", round(.lower, 3),
                      ", ", round(.upper, 3), "]"),
        .rope_overlap = sub("^(-?)0.", "\\1.",
                            sprintf("%.3f", .rope_overlap))) |> 
    select(.variable_name, .median, .lower, .upper, .hdi, .rope_overlap)

tbl_data |> 
    mutate(viz = "") |> 
    select(-c(.lower, .upper)) |> 
    knitr::kable(
        format = "latex", 
        escape = FALSE,
        col.names = c("",
                      "$\\beta$",
                      "95\\% HDI",
                      "$p(\\text{ROPE})$",
                      ""),
        digits = 3,
        booktabs = TRUE,
        align = "lclrc",
        linesep = ""
    ) |> 
    kableExtra::kable_styling() |> 
    kableExtra::pack_rows("Intercepts (at 22 months)",
                          start_row = 1, 
                          end_row = 2) |> 
    kableExtra::pack_rows("Slopes (upper bound)",
                          start_row = 3,
                          end_row = 10) |> 
    kableExtra::column_spec(
        5,
        background = "#f2f2f2",
        image = kableExtra::spec_pointrange(
            x = tbl_data$.median, 
            xmin = tbl_data$.lower, 
            xmax = tbl_data$.upper, 
            vline = 0,
            xaxt = "n",
            col = "#000000",
            res = 200)
    )

>>>>

---
abstract: |
    Bilingual infants' developmental trajectories of lexical acquisition are equivalent to those shown by their monolingual peers. This is remarkable, given the increased complexity of their linguistic input. Recent studies suggested that bilingual vocabulary growth is boosted by the number of cognates shared by the pair of languages being learned, and that this facilitation effect is driven by a stronger parallel activation of cognates during linguistic exposure, compared to non-cognates. The mechanisms behind this facilitation are still unclear. In this study, we capitalised on accumulator models of language acquisition to propose an account of bilingual lexical acquisition in which parallel activation increases the rate at which children accumulate learning instances for words in both languages, even in fully monolingual situations. Under this hypothesis, we predicted a stronger cognate facilitation for words to which children were exposed less frequently (low-exposure words), as they are co-activated by their translation more often than high-exposure words do. We developed an extensive online vocabulary checklist, the Barcelona Vocabulary questionnaire (BVQ), to collect vocabulary data from 366 Catalan-Spanish bilingual toddlers aged 12 to 32 months. We then used Bayesian explanatory item response theory to model the acquisition trajectories of 302 Catalan/Spanish translation equivalents. We found an interaction between exposure and cognateness, which pointed to cognateness facilitating the aquisition of low-exposure words, but not of mean exposure or high-exposure words. Overall, our findings suggest that cognateness plays a key role in bilingual lexical acquisition, and provides evidence for a frequency-mediated facilitation effect driven by parallel activation.

thanks: |
    The authors declare no conflicts of interest with regard to the funding source of this study. GGC was supported by a FPI 2020 research contract (PRE2019-088165) in a project funded by the Spanish Ministry of Science, Innovation, and Universities (PGC2018-101831-B-I00). DAV was supported by the European Union's Horizon 2023 research and innovation program under Marie Skłodowska–Curie Grant (765556) and a postdoctoral fellowship from the Foundation for Science and Technology of Portugal (UIDB/00214/2020). IC was supported by the Investigo program funded by the European Union's NextGenerationEU (NGEU) recovery plan. NSG was supported by Academia 2019 award from the Catalan Institution for Research and Advanced Studies (ICREA). We are grateful to Chiara Santolin, Ege E. Özer, and the rest of the Speech Acquisition and Perception research group, and to Alicia Franco-Martínez and Cristina Rodríguez-Prada, for their helpful feedback. We thank Xavier Mayoral, Silvia Blanch, and Cristina Cuadrado for their technical support, and Cristina Dominguez and Katia Pistrin for their efforts in recruiting infants. We also thank all families and infants who participated in the experiments.
---

@@

options(knitr.kable.NA = "-")

@@

The foundations of word learning are in place early in age. At six months, infants start directing their gaze to some objects when hearing their labels [@tincoff1999beginnings; @bergelson2012months; @bergelson2015early], and shortly after caregivers start reporting some words as acquired by their infant in vocabulary checklists [e.g., @fenson2007macarthurbates; @samuelson2021precision]. Most research on early word acquisition relies extensively on data from monolingual children, and is oblivious to the fact that a substantial proportion of the world population acquires more than one language from early ages [@grosjean2021extent]. Previous work on bilingual vocabulary acquisition pointed to bilingual toddlers knowing, on average, less words in each of their languages than their monolinguals peers, and to both groups knowing a similar number of words---if not more words---when the bilinguals’ two languages are pooled together. @hoff2012dual found that English-Spanish bilingual toddlers in South Florida (United States) knew less words in English than monolinguals did, but both groups knew a similar total amount of words when both English and Spanish vocabularies were counted together. Other studies have provided converging evidence that bilinguals know a similar or even larger number of words than monolinguals when the two languages are aggregated [@oller2002language; @pearson1994patterns; @pearson1993lexical; @patterson2004comparing; @patterson2004bilingual; @smithson2014bilingualism; @petitto2001bilingual; @gonzalez-barrero2020bilingual]. A more detailed analysis of the words in bilinguals’ lexicon shows some interesting patterns.

One important observation of studies on bilinguals' early vocabulary acquisition is that cognate words are easier to acquire than non-cognate words. Cognate words are translation equivalents that are phonologically similar (or share some type of form-similarity). For instance, the Spanish translation equivalent of *cat* is *gato*, a cognate word; the translation equivalent of *dog* is *perro*, a non-cognate word. The differences in the percentages of cognate and non-cognate words between two languages is related to historical reasons: languages typologically close (like Dutch and English or Italian and Spanish) share more cognates than languages typologically distant (like English and Chinese, or Urdu and Spanish). The conclusion that cognate words are easier to learn is based on two types of evidence: studies investigating vocabulary sizes in children learning language pairs with different percentages of cognates (that is, differing in their typological distance) and studies comparing the number of cognate and non-cognate words children know in a specific language pair.

@floccia2018introduction published an impressive study comparing vocabularies of children learning several language pairs differing in their percentage of cognates. The authors collected vocabulary data on word comprehension and production from 372 24-month-old bilingual toddlers living in the United Kingdom who were learning English and an additional language. The additional language was one of 13 typologically diverse languages: Bengali, Cantonese Chinese, Dutch, French, German, Greek, Hindi/Urdu, Italian, Mandarin Chinese, Polish, Portuguese, Spanish, and Welsh. The authors calculated the average lexical overlap between the words in each of these additional languages and their translation equivalents in English, which was taken as a proxy of the degree of cognateness between each pair of languages. Floccia and co-workers reported an increase in vocabulary size in the additional language (i.e., not English) associated with an increase in the average phonological similarity between the translation equivalents of each language pair. For example, English-Dutch bilinguals (languages with a high phonological overlap), were able to produce more Dutch words than English-Mandarin bilinguals (languages with a low phonological overlap) were able to produce in Mandarin. @blom2020crosslanguage, @bosma2019longitudinal, and @gampe2021does reported similar results, providing converging evidence of a facilitatory effect of a lower language distance (i.e., higher degre of cognateness) on vocabulary size.

A second set of studies suggested that cognates are overrepresented in bilinguals’ early lexicon. @bosch2014first collected parental reports of expressive vocabulary from 48 Catalan-Spanish bilinguals aged 18 months and found that cognates represented a larger proportion of participant’s vocabulary than non-cognates. @schelletter2002effect provided converging evidence from a longitudinal single-case study, in which an English-German bilingual child produced cognates earlier than non-cognates, on average. But the high proportion of cognates in the vocabulary of the participants in these two studies may not necessarily evidence of a facilitation effect of cognateness, but rather of simply the high proportion of cognates present in the pair of languages being learned. For instance, if two given languages share a high proportion of cognates like $\sim 70$%, the vocabulary contents of children learning both languages should, in principle, approximate such proportion of cognates, even in the absence of a cognateness facilitation effect. More recently, @mitchell2022cognates addressed this issue in a longitudinal study with a larger sample. The authors collected expressive vocabulary data of 47 16- to 30-month-old French-English bilinguals living in Canada, in both languages. They created two lists of translation equivalents: one made of 131 cognates, and one made of 406 non-cognates. The proportion of words that children were reportedly able to produce was higher in the cognate lists than in the non-cognate list across ages. Critically, this difference persisted after both lists were matched in size, controlling their semantic category (i.e., furniture, animals, food were similarity represented in both lists) and age-of-acquisition norms (an index of word difficulty). Taken together, the results of these two lines of research have been considered as providing support to the hypothesis that phonological similarity (as reflected in cognateness) plays a facilitation role in bilingual word acquisition.

@@

We propose an account in which a learning instance for a word may also represent a learning instance for its translation equivalent, to the extent that such translation equivalent is co-activated^[We use the term *learning instance* in the fashion of accumulator models of language acquisition; as an exposure to a word-form that constitutes an opportunity for the child to accumulate information about the word. We do not make strong assumptions about whether a learning instance is a discrete or a continuous unit of accumulation of information. We rather consider that a learning instance of a word is so to the degree of the strength of activation of its phonological representation. This activation may result from the infant being exposed to the actual word-form, or the result of activation spreading through phonological or semantic links across lexical representation, as in the case of parallel activation.]. The strength of this co-activation is proportional to the phonological similarity between the two translation equivalents; given that cognates share higher phonological similarity than non-cognates, the former should be co-activated more strongly than the latter. This should lead to a faster accumulation of learning instances for cognates, compared to non-cognates. Parallel activation would allow bilingual children to accumulate learning instances for words in both languages even during fully monolingual situations, but the impact of this mechanism would be asymmetric across languages: words from the lower-exposure language would receive stronger activation from words in the higher-exposure language than vice versa. Therefore, the acquisition of words from the lower-exposure language would benefit more strongly from their cognate status than words from the higher-exposure language. This asymmetric cross-language activation would be consistent with previous reports of larger priming effects from the dominant to the non-dominant language,

@@

n_item_language <- bvq_data$pool |> 
    count(language) |>
    pull(n) |>
    purrr::set_names(c("catalan", "spanish"))

n_items_version <- bvq_data$pool |>
    unnest(version) |> 
    count(version, language) |> 
    pivot_wider(names_from = language,
                values_from = n,
                names_repair = janitor::make_clean_names) |> 
    mutate(version = tolower(version))

@@

tbl_data <- items |> 
    distinct(te, .keep_all = TRUE) |> 
    arrange(language, list) |> 
    mutate(a = grepl("A", list),
           b = grepl("B", list),
           c = grepl("C", list),
           d = grepl("D", list)) |> 
    summarise(across(a:d, sum), 
              items = list(meaning),
              .by = semantic_category) |> 
    mutate(
        total = rowSums(across(a:d)),
        items = purrr::map(items, function(x) {
            paste(tail(x, 2),
                  collapse = ", ")
        })) |>  
    unnest_longer(items) |> 
    arrange(desc(total)) |> 
    relocate(semantic_category, a:d, total, items) |> 
    select(-total)

@@

kable(tbl_data,
      format = "latex", 
      digits = 1, 
      booktabs = TRUE,
      col.names = col_names,
      align = "lccccl")  |> 
    row_spec(nrow(tbl_data)-1, 
             hline_after = TRUE)
```

@@

suppressWarnings({
    diff_responses <- responses |>
        distinct(id, age) |>
        left_join(distinct(participants, id, time_stamp, age),
                  by = c("id", "age")) |> 
        group_by() |> 
        arrange(id, time_stamp) |> 
        mutate(diff_lag = time_stamp-dplyr::lag(time_stamp, default = NA),
               .by = id) |>
        summarise(diff_min = min(diff_lag, na.rm = TRUE),
                  diff_max = max(diff_lag, na.rm = TRUE),
                  .by = id) |> 
        dplyr::filter(!(diff_min %in% c(-Inf, Inf))) |>
        summarise(diff_min = min(diff_min, na.rm = TRUE),
                  diff_max = max(diff_max, na.rm = TRUE))
})

@@

dates <- as.Date(range(participants$time_stamp, na.rm = TRUE))

@@

tbl_data <- participants |>
    mutate(across(starts_with("doe_"), 
                  \(x) floor(x * 10) / 10)) |> 
    mutate(
        age = cut(floor(age),
                  c(10, 14, 18, 22, 26, 30, 34, 36),
                  include.lowest = TRUE),
        doe_catalan = as.character(
            cut(doe_catalan,
                c(0, 0.25, 0.5, 0.75, 1),
                labels = c("0-25%",
                           "25-50%", 
                           "50-75%",
                           "75-100%"),
                include.lowest = TRUE
            ))
    ) |>
    count(age, doe_catalan) |>
    arrange(desc(doe_catalan)) |>
    pivot_wider(names_from = age, 
                values_from = n,
                names_repair = janitor::make_clean_names) |> 
    mutate(across(where(is.numeric), \(x) ifelse(is.na(x), 0, x)))

sum_rows <- function(x) if (is.numeric(x)) sum(x) else "N"
sums_row <- as.data.frame(lapply(tbl_data, sum_rows),
                          col.names = colnames(tbl_data))

kable(tbl_data,
      format = "latex", 
      digits = 1, 
      booktabs = TRUE,
      col.names = col_names,
      align = "lcccccc")  |> 
    add_header_above(c(" " = 1,
                       "Age (months)" = ncol(tbl_data)-1)) |>
    row_spec(nrow(tbl_data)-1, 
             hline_after = TRUE)
```

@@

n_items_final <- n_items_duplicated |> 
    count(language) |> 
    group_split(language) |> 
    purrr::set_names(c("Catalan", "Spanish")) |>
    purrr::map_int("n")

@@

rope_interval <- c(lower = -0.1/4, upper = 0.1/4, accuracy = 0.1)

@@

We developed the *Exposure* predictor to account for the fact that bilinguals' exposure to a given word-form is not only a function of the word-form's lexical frequency, but also of the quantitative input they receive from the language such word-form belongs to, we expressed lexical frequencies as the product between both variables. First, we extracted the child-directed lexical frequency of each word from the CHILDES database [@macwhinney2000childes; @sanchez2019childesdb]. Using the corresponding lexical frequencies directly from Catalan and Spanish was not possible due to the low number of Catalan participants and tokens available in their corresponding CHILDES corpora, so they were extracted from the English corpora instead. We mapped the lexical frequencies of the English words to their Catalan and Spanish translations [see @fourtassi2020growth for a similar approach], and transformed them to Zipf scores [@vanheuven2014subtlexuk; @zipf1949human]. We multiplied the resulting lexical frequencies by the reported degree of exposure of the child to Catalan or Spanish. For instance, for a child whose degree of exposure is 80% for Catalan and 20% for Spanish, the expected *Exposure* score to the Catalan word *cotxe* [*car*]---with a lexical frequency of `r round(items$freq[items$item=="cotxe"], 2)`---would be `r round(items$freq[items$item=="cotxe"]*0.80, 2)`, while that of its translation to Spanish *coche* would be `r round(items$freq[items$item=="coche"]*0.20, 2)`.

We defined *Cognateness* as the phonological similarity between each word-form and its translation. For each translation equivalent, we used the `stringdist` [@vanderloo2014stringdist] R package to calculate the Levenshtein distance between the Catalan and the Spanish phonological transcriptions of the word-forms. The Levenshtein distance measures the number of editions (insertions, deletions, or substitutions) that one string of phonemes/characters must go through to become identical to the other [@levenshtein1966binary]. We divided the Levenshtein distance of each translation equivalent by the length of the longest word-form to correct for word length (longer strings are likely to show a larger number of mismatches). Finally, we subtracted the result from one so that it could be interpreted in terms of phonological similarity, instead of phonological distance. This led to a distance metric that ranged from zero to one, where zero indicates that both word-forms are completely different (e.g., /\textipa{"taw.l5}/--/\textipa{"me.sa}/, *table*), and one indicates that the two word-forms are identical (e.g., /\textipa{"mar}/--/\textipa{"mar}/, *sea*) [see @floccia2018introduction; @heeringa2003norwegian; @schepens2012distributions; @fourtassi2020growth; @laing2022phonological for similar approaches].

@@

We implemented the model using `brms` [@burkner2017brms], a R interface to the Stan probabilistic language (`r stan_version`) [@carpenter2017stan]. We ran four iteration chains using the by-default No U-Turn Sampler algorithm with `r format(niterations(model_fit), big.mark = ",")` iterations each and an additional `r format(niterations(model_fit), big.mark = ",")` warm-up iterations per chain. Model posterior draws and predictions were handled using the `tidybayes` [@kay2021tidybayes] R package.

@@

post_draws_list <- model_summary |> 
    mutate(across(.value:.lower,
                  \(x) ifelse(grepl("Intercept", .type),
                              plogis(x), x/4))) |> 
    split(model_summary$.variable)
```

@@

```{r tbl-coefs}
#| label: tbl-coefs
#| tbl-cap: "Posterior distribution of regression coefficients. Median: median of the posterior distribution in the probability scale. 95\\% HDI: 95\\% highest density interval of the distribution. *p*(ROPE): overlap between the 95\\% HDI and the ROPE, indicating the posterior probability that the true value of the coefficient is equivalent to zero. In the right column, we show a graphical representation of the median and 95% HDI of each coefficient, with a reference dotted line indicating the location of zero."
#| echo: false
#| message: false
#| warning: false
#| out-width: 80%
# summarise posterior draws
intercept_label <- "Intercepts (at 22 months)"

tbl_data <- model_summary |>
    select(.variable_name, .value, .lower, .upper, .rope, .type) |>
    mutate(
        across(.value:.upper,
               \(x) ifelse(.type==intercept_label, plogis(x), x/4)),
        .variable_name = gsub("×", '$\\\\times$', .variable_name),
        .hdi = paste0("[", round(.lower, 3),
                      ", ", round(.upper, 3), "]"),
        .rope = ifelse(is.na(.rope),
                       "-",
                       sub("^(-?)0.", "\\1.", sprintf("%.3f", .rope)))
    ) |> 
    select(.variable_name, .value, .lower, .upper, .hdi, .rope)

col_names <- c("", "$\\beta$", "95\\% HDI", "$p(\\text{ROPE})$")

tbl_data |> 
    select(-c(.lower, .upper)) |> 
    kable(format = "latex", 
          escape = FALSE,
          col.names = col_names,
          digits = 3,
          booktabs = TRUE,
          align = "lclr",
          linesep = ""
    ) |> 
    kable_styling() |> 
    pack_rows("Intercepts (at 22 months)",
              start_row = 1, 
              end_row = 2) |> 
    pack_rows("Slopes (upper bound)",
              start_row = 3,
              end_row = 10) 
```
